Off the beaten path: \\ Gaia reveals GD-1 stars outside of the main stream

### 2018-05-01 paper/gd1.tex
trimmy

\begin{abstract}\noindent % trust me
Tidally-disrupted globular clusters are transformed into thin, dynamically-cold
streams of stars that are extremely valuable tracers of the large- and
small-scale distribution of mass in the Galaxy.
Using data from the \gaia\ second data release combined with
\pans\ photometry, we present a sample of highly-probable members of the
longest cold stream in the Milky Way, GD-1.
The resulting map of GD-1: (1) extends the apparent length of the stream by
$20^\circ$, (2) reveals a plausible location for the progenitor, (3) detects
high-contrast gaps along the stream, and (4) indicates
the existence of stream members perturbed off the main stream track.
These discoveries were only possible because of the exquisite proper motions
from \gaia, which permit a clean separation of the stream from the Milky Way
stars.
The additional length and proper treatment of the progenitor will aid in dynamical modeling of GD-1 and map the Galaxy on large scales.
The complex morphology of the stream points to a turbulent history; detailed phase-space properties of the perturbed stream members could potentially constrain dark matter substructure in the Milky Way.

@@

We retrieve \gaia\ data along the previously-identified track of the GD-1 stream
from the \gaia\ science archive\footnote{\url{https://gea.esac.esa.int/}} by
selecting sources with small parallax, $\varpi < 1~\textrm{mas}$ in fields along the GD-1 stream
% , and select
% batches of stars that have small latitude in the heliocentric GD-1 coordinate
% system, $(\phi_1, \phi_2)$, defined in \cite{Koposov:2010}.
We convert the equatorial sky coordinates and proper motions %of the parallax-selected stars
to the GD-1 coordinate system \citep[$\phi_1, \phi_2$,][]{Koposov:2010}, and correct the proper motions for solar reflex
motion by assuming that stars at a given stream longitude, $\phi_1$, have a
distance given by $d(\phi_1) = (0.05 \, \phi_1 + 10)~\textrm{kpc}$;
we assume a solar velocity $\bs{v}_\odot = (11.1, 232.24, 7.25)~\kms$
(\citealt{Schonrich:2010, Bovy:2015}).
\figurename~\ref{fig:selection} (top right) shows stars with $|\phi_2| <
1^\circ$ in solar-reflex-corrected proper motion components in the GD-1 system,
$\mu_{\phi_1, \odot}$ and $\mu_{\phi_2, \odot}$.
GD-1 members are visible as the overdensity around $(\mu_{\phi_1, \odot},
\mu_{\phi_2, \odot}) \approx (-8, 0)~\masyr$.

@@

To improve the contrast of the stream over the background, we cross-match the
sample to the \pans\ photometric catalog. % and use the photometry to clean the sample.
\figurename~\ref{fig:selection} (bottom right) shows the proper-motion-selected
candidate stream stars within $\left|\phi_2\right| < 1^\circ$ in a \pans\
color-magnitude diagram (CMD) de-reddened following \citet{Schlafly:2011}.
GD-1 stellar population stands out from the Milky Way foreground, and is well matched by a 12\,Gyr, $\feh = -1.35$ MIST isochrone at a distance of $7.8\,\kpc$ \citep[red line in Figure~\ref{fig:selection},][]{Dotter:2016, Choi:2016, Paxton:2011}.
% The over-plotted isochrone (red line) is from the MESA Isochrones \& Stellar
% Tracks (MIST; \citealt{Dotter:2016, Choi:2016, Paxton:2011}) and represents a
% $12~\textrm{Gyr}$ old population with $\feh = -1.35$ at a distance of
% $7.8~\kpc$.
We use this isochrone to generate a polygonal selection in de-reddened $g-i$
color and apparent $g$-band magnitude (shaded region in the
bottom right panel of \figurename~\ref{fig:selection}).

@@

Clear density variations, also apparent in the 2D positions of the sources,
manifest as sharp features in the surface density estimates along the stream.
These do not map to features in dust extinction nor are they caused by completeness variations
in the \gaia\ scanning pattern, as 99\% of our sample has at least 8 \gaia\ visibility periods,
promising a robust estimate of the proper motions.
% First, the visible portions of the GD-1 stream are located at high Galactic
% latitudes ($b > 20^\circ$).
%, and we therefore do not expect significant dust
% extinction or variations in extinction within the stream region.
% The median $V$-band extinction over the stream track is
% $\textrm{med}\left(A_V\right) = 0.04~\textrm{mag}$ with a dispersion
% $\sigma_{A_V} \approx 0.03~\textrm{mag}$, as computed from the SFD extinction
% map (\citealt{Schlegel:1998}).
% As the stream approaches the Galactic disk ($\phi_1 < -60^\circ$),
% dust extinction becomes appreciable, and the stream stars are harder to
% select apart from the background.
% Second, 99\% of the stars in our sample have at least 8 \gaia\ visibility
% periods (median and root-variance of 14 and 3, respectively), promising a robust
% estimate of the proper motions.
% The variance in the number density of faint stars originating from the \gaia\
% scanning pattern is only $\approx 1~\textrm{deg}^{-2}$, as estimated using stars
% with proper motions $-9 < \mu_{\phi_1, \odot} < -4.5~\masyr$ and $1 <
% \mu_{\phi_2, \odot} < 4~\masyr$, far smaller than the density variations in the
% stream.
% The relative ratios between different over- and under-densities might be affected by the \gaia\ completeness at the faint end, but they
The observed density variations are therefore likely real morphological changes
along the GD-1 stream.
% as have been suggested before photometric selection alone
% (\citealt{Carlberg:2013, DeBoer:2018}).
The deep under-densities we see from the kinematically-cleaned sample correspond
to previously reported gaps in the stream (\citealt{Carlberg:2013, DeBoer:2018}).
There is also a well-defined, sharp overdensity in the surface density profile
close to $\phi_1 \approx -13^\circ$ with roughly symmetric under-densities on
either side:
this is suggestive of a progenitor system in the final stages of dissolution
(\citealt{Balbinot:2018}).
This region is also the thinnest part of the stream, as would be expected for
recently stripped material, but we defer a robust analysis of variations in the
stream width to future work.
% that has not yet phase-mixed in multiple dimensions.

@@

To initialize the model stream generation, we assume that the present day
position of the progenitor system is at $\phi_1 = -13.5^\circ$, the peak of the
observed surface density profile.
We numerically integrate the backwards from this location for $4~\textrm{Gyr}$,
then forward-generate the model stream, and release stream particles at each
timestep.
We assume an initial progenitor mass of $M=10^5~\msun$, and linearly decrease
the mass of the progenitor until its full disruption at the present.
% day, where we assume $M = 0~\msun$
% (i.e. fully disrupted).
Stream properties computed from this resulting model stream are plotted in
\figurename~\ref{fig:track-and-model} as dashed (orange) lines.
The model stream qualitatively matches the stream track in all phase-space
dimensions, but has a much smoother surface density profile along the extent of
the stream.
The broadening in the observed stream at negative longitudes is also seen in the model stream and is expected for older parts of the stream that have phase-mixed for longer times.
We therefore don't expect that the observed surface density variations can be fully explained
by, e.g., epicyclic overdensities (\citealt{Kupper:2012}).

@@

The central panel of \figurename~\ref{fig:features} shows the map of GD-1, with fields of interest marked in shades of purple for spur-related fields, and shades of blue for blob-related fields.
In both cases, we analyze the stream region (darkest shade), overdensity off from the stream (medium shade) and use a field on the opposite side of the stream from the overdensity as a control (lightest shade).
The $\phi_1$ range is the same for all three fields associated with a given feature, and the control and feature field have the same area, while the stream field is smaller to avoid cross-contamination.

@@

On the other hand, the control fields have some stars entering both the CMD and proper-motion selection boxes, but without any significant overdensities in either.
CMDs of the spur and blob fields have more stars in the selection box than their respective control fields, and the distribution within the box is qualitatively similar to that of the corresponding stream field, suggesting a similar distance.
In proper motions, the spur and blob fields also display properties similar to the adjacent stream fields, with clear overdensities of stars in the selection box.

Stars from the spur and the blob are similar to those along the main stream track in the color-magnitude diagram and in proper motions, which confirms their association to GD-1.
The proximity of these off-stream features to large gaps in the stream is suggestive of the gaps being their origin, further implied by hints of relative velocity differences.
For example, the median proper motion of the blob is offset from the adjacent stream field by $\approx0.3$\,mas\,yr$^{-1}$ in the positive $\mu_{\phi_1}$ direction.
We defer detailed modeling of these features to infer their co-evolution with the stream for a future study.

@@

First, the absence of a clear progenitor has been a long-standing hurdle in dynamical modeling of GD-1, mostly due to the unknown true extent of the stream.
Confirming the progenitor location suggested here, $(\phi_1, \phi_2) = (-13.5^\circ,-0.5^\circ)$, for example by measuring the (canonical) angle variations along the stream \citep{Bovy:2014}, would motivate further searches for stream members along the trailing arm (positive $\phi_1$) to symmetrically match the extent of the leading arm.
The stream length is directly proportional to the information it provides on the underlying gravitational potential \citep{Bonaca:2018}, making GD-1 a top modeling priority to map the inner 30\,kpc of the Galaxy, even at the length of $\sim90^\circ$ inferred in this work.
Properly accounting for the progenitor would also unleash the full constraining power of detailed density structure within the tails, as advocated by \citet{Kupper:2015}.

@@

Confirming the origin of gaps and the off-track features in GD-1 will require
both detailed modeling of the stream formation in the presence of massive
perturbers and follow-up radial velocity measurements of candidate GD-1 members.
Radial velocities will help remove the remaining contamination, further
improving the contrast of the stream, and will enable measurements of energy and
angular momentum along the stream and off-track features.
While constraints on subhalo interactions from individual gaps leave serious
degeneracies between, e.g., the time of the encounter and the mass of the
perturber, energy offsets are time-independent and are a promising avenue
towards placing stronger constraints on the nature of the stream and, if
confirmed, the particular pertuber.

>>>>

\begin{abstract}\noindent % trust me
Tidally-disrupted globular clusters are transformed into thin, dynamically-cold
streams of stars that are extremely valuable tracers of the large- and
small-scale distribution of mass in the Galaxy.
Using data from the \gaia\ second data release combined with
\pans\ photometry, we present a sample of highly-probable members of the
longest cold stream in the Milky Way, GD-1.
The resulting map of GD-1: (1) extends the apparent length of the stream by
$20^\circ$, (2) reveals a plausible location for the progenitor, (3) detects
high-contrast gaps along the stream, and (4) indicates
the existence of stream members perturbed off the main stream track.
These discoveries are only possible because of the exquisite astrometry
from \gaia, which permits a clean separation of the stream from Milky Way
stars.
The additional length and a proper treatment of the progenitor will aid in dynamical modeling of GD-1 for mapping the large-scale dark matter distribution.
The complex morphology of the stream points to a turbulent history; detailed phase-space properties of the perturbed stream members could potentially constrain dark matter substructure in the Milky Way.

@@

We retrieve data along the previously-identified track of the GD-1 stream
from the \gaia\ science archive\footnote{\url{https://gea.esac.esa.int/}} by
selecting sources with small parallax, $\varpi < 1~\textrm{mas}$ in fields along the GD-1 stream
% , and select
% batches of stars that have small latitude in the heliocentric GD-1 coordinate
% system, $(\phi_1, \phi_2)$, defined in \cite{Koposov:2010}.
We convert the equatorial sky coordinates and proper motions %of the parallax-selected stars
to the GD-1 coordinate system \citep[$\phi_1, \phi_2$,][]{Koposov:2010}, and correct the proper motions for solar reflex
motion by assuming that stars at a given stream longitude, $\phi_1$, have a
distance given by $d(\phi_1) = (0.05 \, \phi_1 + 10)~\textrm{kpc}$;
we assume a solar velocity $\bs{v}_\odot = (11.1, 232.24, 7.25)~\kms$
(\citealt{Schonrich:2010, Bovy:2015}).
\figurename~\ref{fig:selection} (top right) shows stars with $|\phi_2| <
1^\circ$ in solar-reflex-corrected proper motion components in the GD-1 system,
$\mu_{\phi_1, \odot}$ and $\mu_{\phi_2, \odot}$.
GD-1 members are visible as the overdensity around $(\mu_{\phi_1, \odot},
\mu_{\phi_2, \odot}) \approx (-8, 0)~\masyr$.

@@

To improve the contrast of the stream over the background, we cross-match the
sample to the \pans\ photometric catalog. % and use the photometry to clean the sample.
\figurename~\ref{fig:selection} (bottom right) shows the proper-motion-selected
candidate stream stars within $\left|\phi_2\right| < 1^\circ$ in a \pans\
color-magnitude diagram (CMD) de-reddened following \citet{Schlafly:2011}.
The GD-1 stellar population stands out from the Milky Way foreground, and is well matched by a 12\,Gyr, $\feh = -1.35$ MIST isochrone at a distance of $7.8\,\kpc$ \citep[red line in \figurename~\ref{fig:selection},][]{Dotter:2016, Choi:2016, Paxton:2011}.
% The over-plotted isochrone (red line) is from the MESA Isochrones \& Stellar
% Tracks (MIST; \citealt{Dotter:2016, Choi:2016, Paxton:2011}) and represents a
% $12~\textrm{Gyr}$ old population with $\feh = -1.35$ at a distance of
% $7.8~\kpc$.
We use this isochrone to generate a polygonal selection in de-reddened $g-i$
color and apparent $g$-band magnitude (shaded region in the
bottom right panel of \figurename~\ref{fig:selection}).

@@

Clear density variations, also apparent in the 2D positions of the sources,
manifest as sharp features in the surface density estimates along the stream.
We have verified that these do not map to features in dust extinction nor are
they caused by completeness variations in the \gaia\ scanning pattern:
the visible portions of the GD-1 stream are located at high Galactic latitudes
($b > 20^\circ$), and 99\% of our sample has at least 8 \gaia\ visibility
periods, promising a robust astrometric solution.
% First, the visible portions of the GD-1 stream are located at high Galactic
% latitudes ($b > 20^\circ$).
%, and we therefore do not expect significant dust
% extinction or variations in extinction within the stream region.
% The median $V$-band extinction over the stream track is
% $\textrm{med}\left(A_V\right) = 0.04~\textrm{mag}$ with a dispersion
% $\sigma_{A_V} \approx 0.03~\textrm{mag}$, as computed from the SFD extinction
% map (\citealt{Schlegel:1998}).
% As the stream approaches the Galactic disk ($\phi_1 < -60^\circ$),
% dust extinction becomes appreciable, and the stream stars are harder to
% select apart from the background.
% Second, 99\% of the stars in our sample have at least 8 \gaia\ visibility
% periods (median and root-variance of 14 and 3, respectively), promising a robust
% estimate of the proper motions.
% The variance in the number density of faint stars originating from the \gaia\
% scanning pattern is only $\approx 1~\textrm{deg}^{-2}$, as estimated using stars
% with proper motions $-9 < \mu_{\phi_1, \odot} < -4.5~\masyr$ and $1 <
% \mu_{\phi_2, \odot} < 4~\masyr$, far smaller than the density variations in the
% stream.
% The relative ratios between different over- and under-densities might be affected by the \gaia\ completeness at the faint end, but they
The observed density variations are therefore likely real morphological changes
along the GD-1 stream.
% as have been suggested before photometric selection alone
% (\citealt{Carlberg:2013, DeBoer:2018}).
The deep under-densities we see from the kinematically-cleaned sample correspond
to previously reported gaps in the stream (\citealt{Carlberg:2013, DeBoer:2018}).
There is also a well-defined, sharp overdensity in the surface density profile
close to $\phi_1 \approx -13^\circ$ with roughly symmetric under-densities on
either side:
this is suggestive of a progenitor system in the final stages of dissolution
(\citealt{Balbinot:2018}).
This region is also the thinnest part of the stream, as would be expected for
recently stripped material; we defer a robust analysis of variations in the
stream width to future work.
% that has not yet phase-mixed in multiple dimensions.

@@

To initialize the model stream generation, we assume that the present day
position of the progenitor system is at $\phi_1 = -13.5^\circ$, the peak of the
observed surface density profile.
We numerically integrate backwards from this location for $4~\textrm{Gyr}$,
then forward-generate the model stream, releasing stream particles at each
timestep.
We assume an initial progenitor mass of $M=10^5~\msun$, and linearly decrease
the mass of the progenitor until its full disruption at the present.
% day, where we assume $M = 0~\msun$
% (i.e. fully disrupted).
Stream properties computed from this resulting model stream are plotted in
\figurename~\ref{fig:track-and-model} as dashed (orange) lines.
The model stream qualitatively matches the stream track in all phase-space
dimensions, but has a much smoother surface density profile along the extent of
the stream.
The broadening in the observed stream at negative longitudes is also seen in the model stream and is expected for older parts of the stream that have phase-mixed for longer times.
We therefore don't expect that the observed surface density variations can be fully explained
by, e.g., epicyclic overdensities (\citealt{Kupper:2012}).

@@

The central panel of \figurename~\ref{fig:features} shows the map of GD-1, with fields of interest marked in shades of purple for spur-related fields, and shades of blue for blob-related fields.
In both cases, we analyze the stream region (darkest shade), overdensity off from the stream (medium shade) and use a field on the opposite side of the stream from the overdensity as a control (lightest shade).
The $\phi_1$ range is the same for all three fields for a given feature, and the control and feature fields have the same area (the stream field is smaller to avoid cross-contamination).

@@

The control fields have some stars in both the CMD and proper-motion selection boxes, but without any significant overdensities in either.
CMDs of the spur and blob fields have more stars in the selection box than their respective control fields, and the distribution within the box is qualitatively similar to that of the corresponding stream field, suggesting a similar distance.
In proper motions, the spur and blob fields also display properties similar to the adjacent stream fields, with clear overdensities of stars in the selection box.

Stars from the spur and the blob are similar to those along the main stream track in the color-magnitude diagram and in proper motions, which confirms their association to GD-1.
The proximity of these off-stream features to large gaps in the stream is suggestive of the gaps being their origin, further implied by hints of relative velocity differences.
For example, the median proper motion of the blob is offset from the adjacent stream field by $\approx0.3$\,mas\,yr$^{-1}$ in the positive $\mu_{\phi_1}$ direction.
We defer detailed modeling of these features to infer their co-evolution with the stream for future work.

@@

First, the absence of a clear progenitor has been a long-standing hurdle in dynamical modeling of GD-1, mostly due to the unknown true extent of the stream.
Confirming the progenitor location suggested here, $(\phi_1, \phi_2) =
(-13.5^\circ,-0.5^\circ)$, would motivate further searches for stream members
along the trailing arm (positive $\phi_1$) to symmetrically match the extent of
the leading arm.
% for example measuring the (canonical) angle variations along the stream \citep{Bovy:2014}
The stream length is directly proportional to the information it provides on the underlying gravitational potential \citep{Bonaca:2018}, making GD-1 a top modeling priority to map the inner 30\,kpc of the Galaxy, even at the length of $\sim90^\circ$ inferred in this work.
Properly accounting for the progenitor would also allow for studying detailed density structure within the tails, as advocated by \citet{Kupper:2015}.

@@

Confirming the origin of gaps and the off-track features in GD-1 will require
both detailed modeling of the stream formation in the presence of massive
perturbers and follow-up radial velocity measurements of candidate GD-1 members.
Radial velocities will help remove the remaining contamination, further
improving the contrast of the stream, and will enable measurements of
differential energy and angular momentum along the stream and off-track
features.
While constraints on subhalo interactions from individual gaps leave serious
degeneracies between, e.g., the time of the encounter and the mass of the
perturber, energy offsets are time-independent and are a promising avenue
towards placing stronger constraints on the nature of the stream and, if
confirmed, the particular pertuber.
